Security vs. Maintainability: \\ Fixing Vulnerabilities Obfuscates your Code

### 2019-01-28 paper/ICPC19/icpc19.tex
merge
several fixes S2

\title{Security vs. Maintainability: Fixing Vulnerabilities Obfuscates your Code}

@@

This paper is structured as follows: section \ref{sec:motivation} introduces an example of a security refactoring of a known vulnerability found in OpenSSL\footnote{Repository details available at https://github.com/openssl/openssl (Accessed on January 27, 2019)} software; section \ref{sec:methodology} describes the methodology used to answer the research questions; section \ref{sec:results} presents the results and section \ref{sec:discussion} discusses their implications; section \ref{sec:threats}, threats to the validity of the study are enumerated; section \ref{sec:rw} describres the different work and existing literature in the field of study; and, finally section \ref{sec:conclusions} concludes the main findings and elaborates on future work.

\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software codebase. However, these refactorings may have a negative
impact on software maintenance, mostly because developers follow the quickest
solution to fix the problem and not necessarily the most elegant one. Thus, we
want to understand how security refactorings impact the maintainability of
software.

One example is the refactoring of the DTLS protocol implementation in OpenSSL after
Adam Langley detect that remote attackers could cause a Denial-of-Service (DoS) attack
when crafting DTLS handshake messages to trigger memory allocations corresponding to large length values.
This vulnerability is listed at the Common Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on January 27, 2019)} and
it is one of the vulnerabilities under evaluation in this study. The impact of this refactoring
on the OpenSSL maintainability was negative. The following snippet presents the codebase changes
performend on the \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636 (Accessed on January 27, 2019)} by the OpenSLL developers to fix the vulnerability.

@@

In the \texttt{dtls1\_reassemble\_fragment} function the value of \texttt{msg\_hdr-$>$frag\_off+frag\_len} was checked against the
maximum handshake message size but \texttt{msg\_len} is allocated to the fragment buffer in line 37. Thus, if the fragment was within
the allowed size, the pending handshake message could consume $16MB$+$2MB$. According with the message commit, 10 handshake messages
are allowed, so the attacker could consume approximately $180MB$ per DTLS connection.

@@

\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to limit the number of bytes allowed in a DTLS handshake message for \texttt{s}. This function replaces the lines 18-21.

@@

\ref{lst:func3} Replaced by attribution in line 23.

@@

Although limiting the size of a handshake message seems an elementary problem to repair, the example shows that there was a significant amount of changes performed in the codebase which produced a negative impact in the project maintainability. Even though the developer tries to simplify the codebase with the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it also disrupts 2 of the guidelines proposed by the Software Improvement Group (SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the \emph{Write Simple Units of Code} guideline. And, it increases the codebase size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which infringes the \emph{Keep Your Codebase Small} guideline.

In this study, our concern is if while improving software security, developers are also inserting a negative impact on the maintainability of their software which may increase technical debt when choosing the easiest solution instead of the using the best approach. To answer the following two research questions, we use a dataset of security refactorings to measure the impact of the changes in the maintainability of open-source software.

@@

There are security flaws that are more difficult to refactor than others. For instance, introducing a mechanism to secure authentication is not so easy to implement as fixing a cross-site scripting vulnerability with a \texttt{escape} function. Knowing which patterns are more likely to increase maintainability issues is one step forward to bring awareness to security engineers of what patterns need more attention. The taxonomy of security patterns used to answer this question is the same as the one provided by the authors of the dataset. Maintainability is measured separately for each pattern.

\section{Methodology}\label{sec:methodology}

In this section, the methodology used to measure the impact of security refactorings on the maintainability of open-source software is presented in depth in the following sections and illustrated in Figure \ref{fig:met}. Aiming to answer the research questions presented in this paper, we use a dataset containing $716$ security refactorings collected from open-source software available on GitHub. $648$ of those refactorings were reported in a previous study conducted in 2017 \cite{Reis:2017:IJSSE}. In parallel, a baseline of regular commits was randomly collected from the list of projects of the main dataset to evaluate what is the impact of regular commits on the maintainability of open-source software. The Software Improvement Group's\footnote{SIG's website is available at https://www.sig.eu/ (Accessed on January 23, 2019)} web-based source code analysis service Better Code Hub (BCH) \footnote{BCH's website is available at https://bettercodehub.com/ (Accessed on January 23, 2019)} was used to collect the maintainability reports for both security and regular commits.

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology \Sof{@todo: Improve it!}}
	\label{fig:met}
\end{figure}

>>>>

\title{Security vs. Maintainability: \\ Fixing Vulnerabilities Obfuscates your Code}

@@

This paper is structured as follows: section \ref{sec:motivation} introduces an
example of a security refactoring of a known vulnerability found in
OpenSSL\footnote{Repository available at
https://github.com/openssl/openssl (Accessed on \today{})} software;
section \ref{sec:methodology} describes the methodology used to answer the
research questions; section \ref{sec:results} presents the results and section
\ref{sec:discussion} discusses their implications; section \ref{sec:threats},
threats to the validity of the study are enumerated; section \ref{sec:rw}
describres the different work and existing literature in the field of study;
and, finally section \ref{sec:conclusions} concludes the main findings and
elaborates on future work.
%
\section{Motivation and Research Questions}\label{sec:motivation}
%
Due to time-to-market pressure and the lack of security expertise, code-related
security flaws are generally detected \textit{after the fact}, i.e., when
hackers exploit them. It turns out that fixing these issues is as simple as
modifying the software codebase. However, these refactorings may have a negative
impact on software maintenance, mostly because developers follow the quickest
solution to fix the problem and not necessarily the most elegant and performant
one. Thus, we want to understand how security refactorings impact the
maintainability of software.

As an example, consider the refactoring of the Datagram Transport Layer Security
(DTLS) protocol implementation in OpenSSL to address the fact that hackers could
cause a Denial-of-Service (DoS) attack by crafting DTLS
handshake messages to trigger memory allocations corresponding to large length
values. This vulnerability is listed at the Common Vulnerabilities and Exposures
dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at
http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on
\today{})}. It is amongst the vulnerabilities studied in our evaluation.
The following snippet presents the changes performend on the
\emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
(Accessed on \today{})} by the OpenSLL developers to fix the vulnerability.

@@

In the \texttt{dtls1\_reassemble\_fragment} function the value of
\texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the maximum
handshake message size, but \texttt{msg\_len} is allocated to the fragment
buffer in line $37$. Thus, if the fragment was within the allowed size, the
pending handshake message could consume $18MB$ ($16MB$+$2MB$). According to the
message commit, $10$ handshake messages are allowed, so the attacker could
consume approximately $180MB$ per DTLS connection.

@@

\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
inserted to limit the number of bytes allowed in a DTLS handshake message for
\texttt{s}. This function replaces lines $18$---$21$.

@@

\ref{lst:func3} Replaced by attribution in line $23$.

@@

Although limiting the size of a handshake message seems an elementary problem to
repair, the example shows that there was a significant amount of changes
performed in the codebase which produced a negative impact in the project
maintainability. Even though the developer tries to simplify the codebase with
the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it
also disrupts $2$ of the guidelines proposed by the Software Improvement Group
(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
\emph{Write Simple Units of Code} guideline. And, it increases the codebase size
in $4$ lines of code ($13$ lines added, $9$ lines deleted) which infringes the
\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
impact of this refactoring on the OpenSSL maintainability was negative.

In this study, our concern is to study whether while improving software
security, developers are also inserting a negative impact on the maintainability
of their software which may increase technical debt when choosing the easiest
solution instead of the using the best approach. To answer the following two
research questions, we use a dataset of security refactorings to measure the
impact of the changes in the maintainability of open-source software.

@@

There are security flaws that are more difficult to refactor than others. For
instance, introducing a mechanism to secure authentication is not so easy to
implement as fixing a cross-site scripting vulnerability with a \texttt{escape}
function. Knowing which patterns are more likely to increase maintainability
issues is one step forward to bring awareness to security engineers of what
patterns need more attention. The taxonomy of security patterns used to answer
this question is the same as the one provided by the authors of the dataset.
Maintainability is measured separately for each pattern.
%
\section{Methodology}\label{sec:methodology}
%
The methodology used to measure the impact of security refactorings on the
maintainability of open-source software is discussed in the following sections
and illustrated in Figure \ref{fig:met}. Aiming to answer the research questions
outlined in the previous section, in this paper, we use a dataset containing
$716$ security refactorings collected from open-source software available on
GitHub. $648$ of those refactorings were reported in a previous study conducted
in 2017 \cite{Reis:2017:IJSSE}. Moreover, a baseline of regular commits was
randomly collected from the list of projects of the main dataset to evaluate
what is the impact of regular commits on the maintainability of open-source
software. The Software Improvement Group's\footnote{SIG's website is available
at https://www.sig.eu/ (Accessed on \today{})} web-based source code
analysis service Better Code Hub (BCH)\footnote{BCH's website is available at
https://bettercodehub.com/ (Accessed on \today{})} was used to quantify
maintainability for both security and regular commits.

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
\end{figure}
